---
title: "Increases in crime-related demand are hitting some forces harder than others"
author: Matt Ashby
date: '2021-07-19'
categories:
  - Crime and justice chart of the week
tags:
  - crime
  - crime pressure
  - crime severity
  - police
---

```{r set knitr options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include=FALSE)
```

```{r set chart parameters}
chart_details <- list(
	id = "crime-pressure-force",
	title = "Increases in crime-related demand are hitting some forces harder than others",
	subtitle = "One way to understand crime-related demands on police forces is to track 'crime pressure', a measure of investigative workload representing the number of crimes reported to a force per officer, weighted according to Office for National Statistics estimates of the relative severity of different crimes. Over the past decade years, this measure has risen in every police force in England and Wales, and has doubled in many places.",
	source_url = "https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/crimeseverityscoreexperimentalstatistics",
	source_title = "Office for National Statistics, 2020"
)
```

```{r load packages and helper}
# custom packages not loaded by helpers.R
library(fable)
library("ggrepel")
library("lubridate")
library("readxl")

# load this after loading custom packages
source(here::here("helpers.R"))
```

```{r get and tidy data}

# get severity score data
if (!file.exists(paste0(chart_details$id, "-data.csv.gz"))) {
  
  # download data
  # Source: https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/crimeseverityscoreexperimentalstatistics
  # data_file <- tempfile(fileext = ".xls")
  data_file <- "~/Downloads/cssdatatool2.xlsx"
  # download.file("https://www.ons.gov.uk/file?uri=%2fpeoplepopulationandcommunity%2fcrimeandjustice%2fdatasets%2fcrimeseverityscoreexperimentalstatistics%2fcurrent/cssdatatool.xls", destfile = data_file)
  
  # load data
  file_data <- read_excel(data_file, sheet = "Data - reg + pfa", skip = 1)
  
  # tidy data
  tidy_data <- file_data %>% 
    # extract only columns relevant to severity scores, not the adjacent table
    # of offence rates
    select(1:21) %>% 
    pivot_longer(c(-`Offence group`, -Code, -Name), names_to = "year", 
                 values_to = "severity_score") %>% 
    mutate(
      # extract the month/year of the end of the data period
      year_ending = str_trim(str_remove(str_extract(year, "\\n.+?\\.{3}"), 
                                        "\\.{3}")),
      # convert this into a date
      year_ending = parse_date(year_ending, format = "%b '%y"),
      # convert date to final day of the month rather than the first
      year_ending = year_ending + days_in_month(year_ending) - days(1)
    ) %>% 
    # select(-year) %>% 
    clean_names()
	
	# save tidy data
	write_csv(tidy_data, paste0(chart_details$id, "-data.csv.gz"))	
	
} else {
	
	# load tidy data
	tidy_data <- read_csv(paste0(chart_details$id, "-data.csv.gz"))
	
}	

# get population data for local authorities, because population data aren't
# published for police force areas
if (!file.exists(paste0(chart_details$id, "-pop-data.csv.gz"))) {
  
  # download data
  # Source: https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland
  pop_data_file <- tempfile(fileext = ".zip")
  download.file("https://www.ons.gov.uk/file?uri=%2fpeoplepopulationandcommunity%2fpopulationandmigration%2fpopulationestimates%2fdatasets%2fpopulationestimatesforukenglandandwalesscotlandandnorthernireland%2fmid2001tomid2019detailedtimeseries/ukdetailedtimeseries20012019.zip", destfile = pop_data_file)
  unzip(
    pop_data_file, 
    files = "MYEB1_detailed_population_estimates_series_UK_(2019).csv", 
    exdir = tempdir()
  )
  
  # load data
  pop_data <- paste0(
    tempdir(), 
    "/MYEB1_detailed_population_estimates_series_UK_(2019).csv"
  ) %>% 
    read_csv() %>% 
    pivot_longer(c(-ladcode19, -laname19, -country, -sex, -age),
                 names_to = "year", values_to = "population", 
                 names_prefix = "population_") %>% 
    count(ladcode19, laname19, year, wt = population, 
          name = "population")
  
  # save tidy data
	write_csv(pop_data, paste0(chart_details$id, "-pop-data.csv.gz"))	
  
} else {
	
	# load tidy data
	pop_data <- read_csv(paste0(chart_details$id, "-pop-data.csv.gz"))
	
}

# load police numbers data
police_numbers <- read_csv("police-uplift-2021-data.csv.gz") %>% 
  # Filter only those rows that relate to counts in March of a year
  filter(month(date) == 3) %>% 
  mutate(
    force_name = case_when(
      force_name == "Avon & Somerset" ~ "Avon and Somerset",
      force_name == "Devon & Cornwall" ~ "Devon and Cornwall",
      force_name == "London, City of" ~ "Metropolitan Police",
      TRUE ~ force_name
    ),
    year = year(date)
  ) %>% 
  # merge MPS and City Police because they're combined in the severity scores
  count(force_name, year, wt = officers, name = "officers")

# get LAD to PFA lookup table
# Source: https://geoportal.statistics.gov.uk/datasets/ons::local-authority-district-to-community-safety-partnerships-to-police-force-areas-december-2019-lookup-in-england-and-wales/about
pfa_lookup <- sf::read_sf("https://opendata.arcgis.com/datasets/5be103b383e047e7abfb6cbe389a9d66_0.geojson") %>%
  sf::st_drop_geometry() %>% 
  clean_names() %>% 
  rename(la_code = lad19cd, pfa_code = pfa19cd, pfa_name = pfa19nm)
```

```{r prepare data for chart}
# get population for police force areas
pfa_population <- pop_data %>% 
  left_join(pfa_lookup, by = c("ladcode19" = "la_code")) %>% 
  # combine MPS and City Police because they're combined in the severity scores
  mutate(
    pfa_code = if_else(pfa_code == "E23000034", "E23000001", pfa_code),
    pfa_name = if_else(pfa_name == "London, City of", "Metropolitan Police",
                       pfa_name)
  ) %>% 
  group_by(pfa_code, pfa_name, year) %>% 
  summarise(population = sum(population)) %>% 
  ungroup() %>% 
  # remove rows that represent Scottish/NI population, for which pfa18cd is NA
  remove_missing(na.rm = TRUE) %>%
  mutate(year = as.numeric(year))

severity <- tidy_data %>% 
  filter(
    offence_group == "Total recorded crime",
    # only severity estimates for police force areas
    str_sub(code, 0, 3) %in% c("E23", "W15"),
    # exclude City Police, since its severity scores are included in the MPS
    name != "City of London",
    # only severity estimates for financial years
    month(year_ending) == 3
  ) %>% 
  select(code, name, year_ending, severity_score) %>% 
  mutate(severity_score = parse_number(severity_score))

severity_total <- severity %>% 
  # create a temporary column containing the year prior to year_ending, for
  # matching to the appropriate *mid calendar year* population estimates
  mutate(year_match = year(year_ending - years(1))) %>% 
  left_join(pfa_population, 
            by = c("code" = "pfa_code", "year_match" = "year")) %>% 
  mutate(year_match = year_match + 1) %>% 
  left_join(police_numbers, 
            by = c("name" = "force_name", "year_match" = "year")) %>% 
  select(name, year_ending, severity_score, population, officers) %>% 
  filter(year_ending >= ymd("2010-01-01")) %>% 
  mutate(total_severity = as.numeric(severity_score) * population) %>% 
  select(name, year_ending, total_severity, officers) %>% 
	mutate(
	  name = str_replace_all(name, " and ", " & "),
		name = str_replace(name, "^North ", "N "),
		name = str_replace(name, "^South ", "S "),
		name = str_replace(name, "^West ", "W "),
		name = str_replace(name, " Yorkshire$", " Yorks"),
		name = recode(name, 
  		"Avon & Somerset" = "Avon & Som",
			"Bedfordshire" = "Beds",
			"Cambridgeshire" = "Cambs",
			"Derbyshire" = "Derbys",
			"Devon & Cornwall" = "Dvn & Corn",
			"Gloucestershire" = "Gloucs",
			"Greater Manchester" = "G Mnchstr",
			"Hampshire" = "Hants",
	  	"Hertfordshire" = "Herts",
			"Lancashire" = "Lancs",
			"Leicestershire" = "Leics",
			"Lincolnshire" = "Lincs",
  		"Metropolitan Police" = "Met Police",
  		"Northamptonshire" = "Northants",
			"Northumbria" = "Northumb",
			"Nottinghamshire" = "Notts",
			"Staffordshire" = "Staffs",
			"Thames Valley" = "Thames V",
			"Warwickshire" = "Warwicks",
			"W Midlands" = "W Mids"
  	)
	)

national_severity <- severity_total %>% 
  group_by(year_ending) %>% 
  summarise_if(is.numeric, sum, na.rm = TRUE) %>% 
  mutate(national_severity_per_officer = total_severity / officers) %>% 
  select(year_ending, national_severity_per_officer)

# calculate severity index
severity_per_officer <- severity_total %>% 
  left_join(national_severity, by = "year_ending") %>% 
  mutate(
    severity_per_officer = total_severity / officers,
    severity_index = severity_per_officer / first(national_severity_per_officer),
    national_severity_index = national_severity_per_officer / first(national_severity_per_officer)
  ) %>% 
  group_by(name) %>% 
  mutate(
    # Change in force-level crime pressure for each year
    change = perc_change(
      first(severity_per_officer), 
      severity_per_officer, 
      format = FALSE
    ),
    # Difference between national and force-level crime pressure
    diff = perc_change(
      national_severity_per_officer,
      severity_per_officer,
      format = FALSE
    )
  ) %>% 
  ungroup()

# order forces by most-recent severity index
forces_in_order <- severity_per_officer %>% 
  filter(year(year_ending) == 2020) %>% 
  arrange(desc(severity_index)) %>% 
  pluck("name")
chart_data <- severity_per_officer %>% 
  group_by(name) %>%
  mutate(
    # Forces must be put in order before setting labels, since the code to 
    # generate the labels relies on the forces already being in order
    name = factor(name, levels = forces_in_order),
    category = factor(case_when(
      last(diff) > 0.1 ~ "higher",
      last(diff) < -0.1 ~ "lower",
      between(last(diff), -0.1, 0.1) ~ "similar",
      TRUE ~ NA_character_
    ), levels = c("higher", "similar", "lower")),
    label = case_when(
      year_ending == last(year_ending) & name == first(forces_in_order) & change > 0 ~ 
        str_glue("2010–20<br>change:<br>+{scales::percent(abs(change), accuracy = 1)}"),
      year_ending == last(year_ending) & name == first(forces_in_order) ~ 
        str_glue("2010–20<br>change:<br>–{scales::percent(abs(change), accuracy = 1)}"),
      year_ending == last(year_ending) & change > 0 ~
        str_glue("+{scales::percent(abs(change), accuracy = 1)}"),
      year_ending == last(year_ending) & is.numeric(change) ~ 
        str_glue("–{scales::percent(abs(change), accuracy = 1)}"),
      TRUE ~ NA_character_
    )
    # label = ifelse(
    #   year_ending == last(year_ending), 
    #   ifelse(
    #     change > 0,
    #     str_glue("+{scales::percent(abs(change), accuracy = 1)}"),
    #     str_glue("–{scales::percent(abs(change), accuracy = 1)}")
    #   ),
    #   NA_character_
    # )
  ) %>% 
  ungroup() %>%
  filter(name != "G Mnchstr")
```

```{r build plot}
chart <- ggplot(chart_data) +
  geom_hline(aes(yintercept = 1), size = 0.25) +
  ggtext::geom_richtext(
    aes(x = min(year_ending), y = max(severity_index), label = label),
    na.rm = TRUE,
    colour = "grey10",
    fill = "white",
    hjust = 0,
    label.padding = unit(0.15, "lines"),
    label.size = NA,
    lineheight = 1,
    size = elements$label_text_size * 0.9,
    vjust = 1
  ) +
  geom_ribbon(
    aes(x = year_ending, ymin = severity_index, ymax = national_severity_index, fill = category),
    alpha = 0.5
  ) +
  geom_line(
    aes(x = year_ending, y = national_severity_index),
    linetype = "21"
  ) +
  geom_line(
    aes(x = year_ending, y = severity_index, colour = category)
  ) +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 1, prefix = "x"),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_colour_manual(
    values = c(
      "higher" = ucl_colours_list[["Light Red"]], 
      "lower" = ucl_colours_list[["Bright Green"]], 
      "similar" = ucl_colours_list[["Light Blue"]]
    ),
    labels = c(
      "higher" = "more than 10% higher   ", 
      "lower" = "more than 10% lower   ", 
      "similar" = "similar (+/-10%)   "
    ),
    guide = guide_legend(title.hjust = 0, title.position = "top"),
    aesthetics = c("colour", "fill")
  ) +
  facet_wrap(vars(name)) +
  labs(
		title = chart_details$title,
		subtitle = format_subtitle(chart_details$subtitle, .width = 120),
    caption = "data for City of London Police and Greater Manchester Police are not available",
    x = NULL,
    y = "crime pressure relative to the 2010 national average",
    colour = "compared to the national average (the dotted line), crime pressure in this force in 2020 was …",
    fill = "compared to the national average (the dotted line), crime pressure in this force in 2020 was …"
  ) +
  theme_cjcharts() +
  theme(
		axis.text = element_text(size = 7),
		legend.key.width = unit(12, "points"), 
		legend.justification = "left",
		panel.spacing = unit(0.2, "lines"),
		strip.text = element_text(size = 8),
		strip.background = element_blank()
  )
```


`r chart_details$subtitle`

```{r display plot, echo=FALSE, include=TRUE, fig.asp=1}
add_logo(chart + labs(title = NULL, subtitle = NULL), 
				 chart_details$source_title, chart_details$id)
```

[larger image](../`r chart_details$id`.png)
| [annotated R code to produce this chart](https://github.com/mpjashby/lesscrime.info/blob/master/content/post/`r chart_details$id`.Rmd)

Data source: [`r chart_details$source_title`](`r chart_details$source_url`)

```{r, echo=FALSE, include=TRUE}
chart_data %>% 
  group_by(year_ending) %>% 
  summarise(across(c("total_severity", "officers"), sum), .groups = "drop") %>% 
  mutate(
    name = "England & Wales", 
    severity_per_officer = total_severity / officers
  ) %>% 
  select(year_ending, name, severity_per_officer) %>% 
  bind_rows(select(chart_data, name, year_ending, severity_per_officer)) %>% 
  mutate(year_ending = year(year_ending)) %>% 
  pivot_wider(names_from = year_ending, values_from = severity_per_officer) %>% 
  left_join(
    select(filter(chart_data, year_ending == last(year_ending)), name),
    by = "name"
  ) %>% 
  group_by(name) %>% 
  mutate(
    label = perc_change(first(`2010`), last(`2020`)),
    across(where(is.numeric), ~ scales::number(. / 1000, accuracy = 0.1)),
    order = as.numeric(`2020`)
  ) %>% 
  ungroup() %>% 
  arrange(desc(order)) %>% 
  select(-order) %>% 
  rename(`change, 2010–20` = label) %>% 
  gt::gt() %>% 
  gt::tab_header(
    title = "Crime severity per officer"
  ) %>% 
  gt::cols_align(align = "right") %>% 
  gt::cols_align(align = "left", columns = name) %>% 
  gt::tab_style(
    style = list(gt::cell_text(weight = "bold")), 
    locations = gt::cells_body(rows = name == "England & Wales")
  )
```

### Notes

The Office for National Statistics (ONS) publishes [estimates of the severity of different types of police-recorded crime](https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/crimeseverityscoreexperimentalstatistics), derived from the average sentence imposed in court for each type of offence. In the same dataset, ONS publishes an overall estimate of the severity of crime per capita in each police force area. To convert this to an estimate of crime severity per officer (referred to above as _crime pressure_), severity per capita was multiplied by police force area population[^fn1] and then divided by the [number of full-time equivalent police officers in each force](https://www.gov.uk/government/collections/police-workforce-england-and-wales). Crime pressure is therefore defined as the severity of police-recorded crime per full-time equivalent officer.

Estimates of the severity of different crimes are inherently limited, although they are likely to be more informative for some purposes than simply counting crimes (which implicitly assumes all crimes are of equal importance). For a further discussion of this, see [_Comparing methods for measuring crime harm/severity_](/publication/measuring_crime_harm/).

[^fn1]: Police force area populations were calculated by aggregating [mid-year population estimates for local authorities](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland) to police force areas using the ONS open geography portal [local authority to police force lookup table](https://geoportal.statistics.gov.uk/datasets/local-authority-district-to-community-safety-partnerships-to-police-force-areas-december-2018-lookup-in-england-and-wales).


```{r export chart}
# save PNG for social media
ggsave(
	filename = paste0(chart_details$id, ".png"), 
	plot = add_logo(chart, chart_details$source_title, chart_details$id), 
	device = "png", 
	width = 600 / 72, 
	height = 600 / 72, 
	units = "in", 
	bg = "white"
)
```
