---
title: "Chart: One in ten of us is on the National DNA Database"
author: Matt Ashby
date: '2019-12-16'
slug: dna-database
categories:
  - Crime and justice chart of the week
tags:
  - DNA
  - criminal justice
  - ethnicity
---

```{r set knitr options, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, include=FALSE)
```

```{r set chart parameters}
chart_details <- list(
	id = "dna-database",
	title = "One in ten of us is on the National DNA Database",
	subtitle = "About 5.3 million people in England and Wales (about 10% of the population aged 10 or over) have their DNA stored on the National DNA Database, one of the largest DNA databases in the world. Police routinely collect DNA from arrested suspects, and it can be stored (often permanently) if they are convicted of almost any offence.",
	source_url = "https://www.gov.uk/government/statistics/national-dna-database-statistics",
	source_title = "Home Office, 2019"
)
```

```{r load packages and helper}
# custom packages not loaded by helpers.R
library("reticulate")

# load this after loading custom packages
source(here::here("helpers.R"))
```

```{r get and tidy data}
if (!file.exists(paste0(chart_details$id, "-data.csv.gz"))) {
	
	# download most-recent data
	new_data_file <- tempfile(fileext = ".ods")
	csv_file <- paste0(new_data_file, ".csv")
	GET("https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/818992/NDNAD_Website_statistics_Q1_19-20.ods", write_disk(new_data_file))
	
	# convert new data from ODS to CSV and read
	source_python(here::here("ods_to_csv.py"))
	ods_to_csv(new_data_file, "Sheet1", csv_file)

	# read data
	file_data <- read_csv(csv_file, skip = 71)
	
	# tidy data
	tidy_data <- file_data %>% 
		slice(1:7) %>% 
		select(1:2) %>% 
		janitor::clean_names() %>% 
		mutate(subject_profiles = as.numeric(subject_profiles))
	
	# save tidy data
	write_csv(tidy_data, paste0(chart_details$id, "-data.csv.gz"))	
	
} else {
	
	# load tidy data
	tidy_data <- read_csv(paste0(chart_details$id, "-data.csv.gz"))
	
}

if (!file.exists(paste0(chart_details$id, "-ethnicity-data.csv.gz"))) {

	pop_data <- read_csv("https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/downloads/age-groups.csv") %>% 
		pivot_longer(-Ethnicity, names_to = "age_group", values_to = "people") %>% 
		janitor::clean_names() %>% 
		write_csv(paste0(chart_details$id, "-ethnicity-data.csv.gz"))
	
} else {
	
	# load population data
	pop_data <- read_csv(paste0(chart_details$id, "-ethnicity-data.csv.gz"))
	
}
```

```{r prepare data for chart}
# harmonise population ethnicity categories
pop_data <- pop_data %>% 
		mutate(ethnicity = case_when(
			str_detect(ethnicity, "White and Black") ~ 
				"Black/African/Caribbean/Black British: Total",
			str_detect(ethnicity, "White and Asian") ~ "Asian/Asian British: Total",
			str_detect(ethnicity, "Other Mixed") ~ 
				"Other ethnic group: Total",
			TRUE ~ ethnicity
		)) %>% 
		filter(
			str_detect(ethnicity, "Total$"), 
			!age_group %in% c("All ages", "Age 0 to 4", "Age 5 to 7", "Age 8 to 9")
		) %>% 
		group_by(ethnicity) %>% 
		summarise(people = sum(people)) %>% 
		filter(ethnicity != "Mixed/multiple ethnic group: Total") %>% 
		mutate(ethnicity = recode(
			ethnicity, 
			"Asian/Asian British: Total" = "Asian", 
			"Black/African/Caribbean/Black British: Total" = "Black", 
			"Other ethnic group: Total" = "Other", 
			"White: Total" = "White"
		)) %>% 
	rename(population = people)

chart_data <- tidy_data %>% 
	mutate(ethnicity = case_when(
		ethnic_appearance %in% c("White North European", "White South European") ~ 
			"White",
		ethnic_appearance %in% c("Chinese, Japanese or SE Asian", 
														 "Middle Eastern") ~ "Other",
		TRUE ~ ethnic_appearance
	)) %>% 
	group_by(ethnicity) %>% 
	summarise(profiles = sum(subject_profiles)) %>% 
	left_join(pop_data, by = "ethnicity") %>% 
	drop_na() %>% 
	mutate(
		# decrease the number of profiles in each category by 14.1% to account for
		# the estimated proportion of DNA profiles that are duplicates
		profiles = profiles * (1 - 0.141),
		not_on_db = population - profiles
	) %>% 
	select(-population) %>% 
	gather("on_ndnad", "people", -ethnicity) %>% 
	mutate(
		ethnicity = factor(str_to_lower(ethnicity), 
											 levels = c("white", "asian", "black", "other")),
		ethnicity = fct_recode(ethnicity, "Asian" = "asian"),
		on_ndnad = on_ndnad == "profiles"
	) %>% 
	arrange(desc(ethnicity)) %>% 
	group_by(ethnicity) %>% 
	mutate(
		on_ndnad = fct_rev(factor(on_ndnad)),
		population = sum(people),
		prop = people / population,
		xmin = ifelse(on_ndnad == TRUE, 0, 1 - prop),
		xmax = ifelse(on_ndnad == TRUE, prop, 1)
	) %>% 
	ungroup() %>% 
	group_by(on_ndnad) %>% 
	mutate(
		ymax = cumsum(population) + (sum(population) * 0.01),
		ymin = ymax - population,
		ycen = ymin + (0.5 * (ymax - ymin)),
		ymax = ymax - (sum(population) * 0.01),
		label = ifelse(on_ndnad == TRUE, as.character(ethnicity), NA)
	) %>% 
	ungroup()

# add chart labels
chart_labels <- tribble(
	~x, ~y, ~xend, ~yend, ~label, ~hjust, ~vjust, ~curve,
	0.1, 56075912 * 0.5, 0.15, 56075912 * 0.5, balance_lines("most people on the DNA database are white, simply because most of the population is white", 3), "left", "center", "right",
	0.13, 56075912 * 0.07, 0.18, 56075912 * 0.18, balance_lines("black people are twice as likely to have their DNA stored on the database, compared to white people", 3), "left", "bottom", "left"
)
```

```{r build plot}
chart <- ggplot(chart_data, aes(xmin = xmin, ymin = ymin, xmax = xmax, ymax = ymax, 
											 fill = on_ndnad)) +
	geom_rect() +
	geom_text(aes(x = 1, y = ycen, label = label), hjust = 0, nudge_x = 0.01,
						na.rm = TRUE, size = (11 * 0.8) / (14 / 5), colour = "grey30") +
	# add explanatory labels
	geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "right"), inherit.aes = FALSE, 
						 curvature = elements$label_line_curvature, 
						 colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_segment(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "straight"), 
						 inherit.aes = FALSE, colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "left"), inherit.aes = FALSE, 
						 curvature = elements$label_line_curvature * -1, 
						 colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_label(aes(x = xend, y = yend, label = label, hjust = hjust, 
								 vjust = vjust),
						data = chart_labels, inherit.aes = FALSE, 
						colour = "black",
						fill = ucl_colours_list[["Light Green"]], 
						size = elements$label_text_size, 
						lineheight = elements$label_text_lineheight,
						label.size = NA, show.legend = FALSE) +
	# end of explanatory labels
	scale_x_continuous(breaks = seq(0, 1, by = 0.1), 
										 labels = scales::percent_format(accuracy = 1), 
										 expand = expand_scale(mult = c(0, 0.05))) +
	scale_y_continuous(expand = c(0, 0), 
										 labels = scales::number_format(accuracy = 1, scale = 1/1e6, 
										 															 suffix = "m")) +
	# the false value is a 90% tint from https://www.color-hex.com/color/500778
  scale_fill_manual(
    values = c(`TRUE` = ucl_colours_list[["Dark Green"]], 
    					 `FALSE` = ucl_colours_list[["Light Green"]]),
    labels = c(`TRUE` = "people on DNA database  ", 
    					 `FALSE` = "people not on DNA database")
  ) +
	coord_cartesian(clip = "off") +
	labs(
		title = chart_details$title,
		subtitle = format_subtitle(chart_details$subtitle),
		legend = NULL,
		x = "proportion of people in an ethnic group",
		y = "population",
		fill = NULL
	) +
	theme_cjcharts() +
	theme(
		panel.grid = element_blank()
	)
```

`r chart_details$subtitle`

```{r display plot, echo=FALSE, include=TRUE}
add_logo(chart + labs(title = NULL, subtitle = NULL), 
				 chart_details$source_title, chart_details$id)
```

[larger image](../`r chart_details$id`.png)
| [annotated R code to produce this chart](https://github.com/mpjashby/lesscrime.info/blob/master/content/post/`r chart_details$id`.Rmd)

Data source: [`r chart_details$source_title`](`r chart_details$source_url`)

The estimated proportion of people on the National DNA Database attempts to take 
account of the presence of duplicate records on the database by reducing the 
number of samples in each category by 14.1%, the estimated proportion of records 
that are duplicates. See the data source for more information.

Since the National DNA Database does not have an equivalent ethnicity category
to the 'Mixed/multiple ethnic group' categories in population data, people in 
these groups have been categorised on this chart as being of Black, Asian or 
other ethnicity.

```{r export chart}
# save PNG for social media
ggsave(
	filename = paste0(chart_details$id, ".png"), 
	plot = add_logo(chart, chart_details$source_title, chart_details$id), 
	device = "png", width = 600 / 72, height = 400 / 72, units = "in"
)
```
