---
title: "Chart: Seizures of the most harmful drugs at UK borders are falling"
author: Matt Ashby
date: '2019-12-02'
slug: drug-seizures
categories:
  - Crime and justice chart of the week
tags:
  - drugs
  - crime
---

```{r set knitr options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include=FALSE)
```

```{r set chart parameters}
chart_details <- list(
	id = "drug-seizures",
	title = "Seizures of the most harmful drugs at UK borders are falling",
	subtitle = "Border Force is responsible for seizing drugs at ports, airports and international parcels depots. However, reports by the Chief Inspector of Borders and Immigration suggest drug-detection efforts are hindered by understaffing, lack of training and focus on conducting passport checks on travellers. Most drug seizures are of small amounts: only 19% of seizures involve more than one kilogram of drugs.",
	source_url = "https://www.gov.uk/government/statistics/seizures-of-drugs-in-england-and-wales-financial-year-ending-2018",
	source_title = "Home Office, 2018"
)
```

```{r load packages and helper}
# custom packages not loaded by helpers.R
library("lubridate")
library("reticulate")

# load this after loading custom packages
source(here::here("helpers.R"))
```

```{r get and tidy data}
if (!file.exists(paste0(chart_details$id, "-data.csv.gz"))) {
	
	# download most-recent data
	new_data_file <- tempfile(fileext = ".ods")
	csv_file <- paste0(new_data_file, ".csv")
	GET("https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/754678/seizures-drugs-mar2018-hosb2618-tables.ods", write_disk(new_data_file))
	
	# convert new data from ODS to CSV and read
	source_python(here::here("ods_to_csv.py"))
	ods_to_csv(new_data_file, "Summary_Table_1", csv_file)

	# read data
	file_data <- read_csv(csv_file, skip = 38, na = c("", ".."))
	
	# tidy data
	tidy_data <- file_data %>% 
		slice(1:28) %>% 
		select(2:14) %>% 
		rename(drug = X2) %>% 
		drop_na() %>% 
		gather("year", "seizures", -drug) %>% 
		mutate(
			drug = str_remove(drug, "[\\d\\s,]+?$"),
			seizures = as.numeric(seizures),
			year_ending = ymd(paste(str_sub(year, -2), "03", "31"))
		) %>% 
		select(year_ending, drug, seizures)

	# save tidy data
	write_csv(tidy_data, paste0(chart_details$id, "-data.csv.gz"))	
	
} else {
	
	# load tidy data
	tidy_data <- read_csv(paste0(chart_details$id, "-data.csv.gz"))
	
}
```

```{r prepare data for chart}
chart_data <- tidy_data %>% 
	filter(
		drug %in% c("Cocaine", "Ecstasy", "Heroin", "Anabolic steroids", 
								"Amphetamines", "Ketamine", "Total cannabis")
	) %>% 
	mutate(
		drug = recode(str_to_lower(drug), "anabolic steroids" = "steroids", 
									"total cannabis" = "cannabis"),
		harm = recode(drug, "heroin" = 1, "cocaine" = 4, "amphetamines" = 5, 
									"cannabis" = 6,"ketamine" = 9, "steroids" = 14, 
									"ecstasy" = 15),
		drug = paste0(drug, " (", scales::ordinal(harm), ")"),
		drug = fct_reorder(drug, harm),
		seizures = seizures / (365.25 / 7)
	)

# add chart labels
chart_labels <- tribble(
	~x, ~y, ~xend, ~yend, ~drug, ~label, ~hjust, ~vjust, ~curve,
	ymd("2011-01-01"), 4, ymd("2011-01-01"), 50, "heroin (1st)", balance_lines("heroin is the most-harmful illegal drug and is only grown overseas, but is rarely seized at the border", 3), "left", "bottom", "left",
	ymd("2018-04-01"), 8, ymd("2017-01-01"), 28, "cocaine (4th)", balance_lines("cocaine seizures by Border Force have fallen by 72% since 2009 even though cocaine use has remained steady", 5), "right", "bottom", "right",
	ymd("2018-04-01"), 9, ymd("2015-04-01"), 19, "ecstasy (15th)", balance_lines("since 2009, ecstasy seizures have increased more than 15-fold", 2), "right", "bottom", "right",
	ymd("2012-10-01"), 74, ymd("2015-10-01"), 64, "ketamine (9th)", balance_lines("rank of relative harm caused by different illegal drugs, according to Nutt et al. 2010", 3), "left", "top", "right",
	ymd("2007-03-31"), 70, ymd("2009-03-31"), 70, "heroin (1st)", "most harmful drugs", "left", "center", "straight",
	ymd("2018-03-31"), 70, ymd("2016-03-31"), 70, "ecstasy (15th)", "least harmful drugs", "right", "center", "straight",
	# ymd(""), 60, ymd(""), 70, "", balance_lines("", 3), "left", "center", "right",
) %>% 
	# order factor levels according to order in chart data, to maintain facet and
	# legend order
	mutate(drug = factor(drug, levels = levels(chart_data$drug)))
```

```{r build plot}
chart <- ggplot(chart_data, aes(year_ending, seizures, fill = drug)) +
	geom_col() +
	# add explanatory labels
	geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "right"), inherit.aes = FALSE, 
						 curvature = elements$label_line_curvature, 
						 colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_segment(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "straight"), 
						 inherit.aes = FALSE, colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "left"), inherit.aes = FALSE, 
						 curvature = elements$label_line_curvature * -1, 
						 colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_label(aes(x = xend, y = yend, label = label, hjust = hjust, 
								 vjust = vjust),
						data = chart_labels, inherit.aes = FALSE, 
						colour = elements$label_text_colour,
						fill = elements$label_text_fill, size = elements$label_text_size, 
						lineheight = elements$label_text_lineheight,
						label.size = NA, show.legend = FALSE) +
	# end of explanatory labels
	scale_x_date(date_breaks = "3 years", date_labels = "'%y") +
	scale_y_continuous(expand = c(0, 0), labels = scales::comma_format()) +
  scale_fill_manual(
    values = unname(ucl_colours_list[c("Orange", "Bright Blue", "Bright Pink",
                                       "Bright Green", "Bright Red", 
                                       "Light Blue", "Yellow")])
  ) +
	coord_cartesian(clip = "off") +
	facet_grid(cols = vars(drug), labeller = label_wrap_gen(width = 10)) +
	labs(
		title = chart_details$title,
		subtitle = format_subtitle(chart_details$subtitle),
		legend = NULL,
		x = NULL,
		y = "average weekly number of seizures by Border Force officers",
		fill = NULL
	) +
	theme_cjcharts() +
	theme(
		legend.position = "none",
		panel.grid = element_blank()
	)
```

`r chart_details$subtitle`

```{r display plot, echo=FALSE, include=TRUE}
add_logo(chart + labs(title = NULL, subtitle = NULL), 
				 chart_details$source_title, chart_details$id)
```

[larger image](../`r chart_details$id`.png)
| [annotated R code to produce this chart](https://github.com/mpjashby/lesscrime.info/blob/master/content/post/`r chart_details$id`.Rmd)

Data source: [`r chart_details$source_title`](`r chart_details$source_url`)

For estimates of the relative harmfulness of different controlled drugs, see
'[Drug harms in the UK: a multicriteria decision analysis](https://doi.org/10.1016/S0140-6736(10)61462-6)' by Nutt et al.

Reports on Border Force can be found on the website of the [Independent Chief Inspector of Borders and Immigration](https://www.gov.uk/government/organisations/independent-chief-inspector-of-borders-and-immigration).

```{r export chart}
# save PNG for social media
ggsave(
	filename = paste0(chart_details$id, ".png"), 
	plot = add_logo(chart, chart_details$source_title, chart_details$id), 
	device = "png", width = 600 / 72, height = 400 / 72, units = "in"
)
```
