---
title: "Chart: Much of the homicide drop has already been erased"
author: Matt Ashby
date: '2019-10-21'
slug: homicide-trends
categories:
  - Crime and justice chart of the week
tags:
  - crime
  - homicide
  - trends
---

```{r set knitr options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include=FALSE)
```

```{r set chart parameters}
chart_details <- list(
	id = "homicide-trends",
	title = "Much of the homicide drop has already been erased",
	subtitle = "The rapid decrease in the homicide (murder, manslaughter and infanticide) rate in England and Wales between 2000 and 2014 was unprecedented over the previous century, but a third of that decrease has already been reversed.",
	source_url = "https://www.gov.uk/government/statistics/historical-crime-data",
	source_title = "Home Office, 2019"
)
```

```{r load packages and helper}
# custom packages not loaded by helpers.R
library("lubridate")
library("readxl")

# load this after loading custom packages
source(here::here("helpers.R"))
```

```{r get and tidy data}
if (!file.exists(paste0(chart_details$id, "-data.csv.gz"))) {
  
  # download data up to 2001 from the Home Office
  data_file <- tempfile()
  download.file("https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/116649/rec-crime-1898-2002.xls", destfile = data_file)
  file_data <- read_excel(data_file, skip = 5)
  
  # download data for 2002 onwards from ONS
  download.file("https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/crimeandjustice/datasets/crimeinenglandandwalesappendixtables/yearendingjune2019/appendixtableyearendingjune201915102019170156.xlsx", destfile = data_file)
  file_data_new <- read_excel(data_file, sheet = "Table A4", skip = 3)
  
  tidy_data <- file_data %>% 
    select(2:3) %>% 
    slice(2:101, 104:107) %>% 
    set_names(c("year", "homicide_count")) %>% 
    mutate(
      year = ymd(paste(str_sub(year, 0, 4), "01", "01")),
      year = if_else(row_number() >= 101, year + months(3), year)
    )
  
  tidy_data_new <- file_data_new %>% 
    select(3:20) %>%
    slice(9) %>% 
    pivot_longer(-`...3`, names_to = "year", values_to = "homicide_count") %>% 
    select(-`...3`) %>% 
    mutate(year = ymd(paste(str_sub(year, 6, 7), "04", "01")))
  
  tidy_data <- bind_rows(tidy_data, tidy_data_new) %>% 
    mutate(homicide_count = as.numeric(homicide_count))
  
	# save tidy data
	write_csv(tidy_data, paste0(chart_details$id, "-data.csv.gz"))	
	
} else {
  
	# load tidy data
	tidy_data <- read_csv(paste0(chart_details$id, "-data.csv.gz"))
  
}	

# get population data
if (!file.exists(paste0(chart_details$id, "-pop-data.csv.gz"))) {
  
  pop_file <- tempfile()
  download.file("https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland/mid2016detailedtimeseries/ukandregionalpopulationestimates1838to2016.zip", destfile = pop_file)
  unzip(pop_file, files = "UK Population Estimates 1838-2016.xlsx", 
        exdir = tempdir())
  
  pop_data <- read_excel(
    paste0(tempdir(), "/UK Population Estimates 1838-2016.xlsx"), 
    sheet = "Table 7", skip = 4
  ) %>% 
    select(Year, Persons) %>% 
    slice(1:179) %>% 
    clean_names() %>% 
    mutate(year = as.numeric(str_remove(year, "Mid-")))
  
  pop_data_new <- read_csv("https://www.ons.gov.uk/generator?format=csv&uri=/peoplepopulationandcommunity/populationandmigration/populationestimates/timeseries/ewpop/pop") %>% 
    slice(8:55) %>% 
    clean_names() %>% 
    set_names(c("year", "persons")) %>% 
    filter(year > 2016) %>% 
    mutate_all(as.numeric)
  
  pop_data <- pop_data %>% 
    bind_rows(pop_data_new) %>% 
    arrange(year)
  
	# save tidy data
	write_csv(pop_data, paste0(chart_details$id, "-pop-data.csv.gz"))	
	
} else {
  
	# load tidy data
	pop_data <- read_csv(paste0(chart_details$id, "-pop-data.csv.gz"))
  
}
```

```{r prepare data for chart}
chart_data <- tidy_data %>% 
  mutate(year = year(year)) %>% 
  filter(year > last(year) - 99) %>% 
  select(year, homicides = homicide_count) %>%
  # move Hillsborough deaths from 2016 to 1989
  mutate(homicides = case_when(
    year == 1989 ~ homicides + 96, 
    year == 2016 ~ homicides - 96, 
    TRUE ~ homicides
  )) %>% 
  left_join(pop_data, by = "year") %>% 
  mutate(
    homicide_rate = (homicides / persons) * 1000000,
    year = ymd(paste(year, "01", "01"))
  )

# add chart labels
chart_labels <- tribble(
	~x, ~y, ~xend, ~yend, ~label, ~hjust, ~vjust, ~curve,
	ymd("1941-01-01"), 9.25, ymd("1937-01-01"), 11, balance_lines("the homicide rate increased during the Second World War", 4), "right", "center", "right",
	ymd("1961-06-01"), 5.5, ymd("1966-06-01"), 4.5, balance_lines("the homicide rate was lowest in 1961", 2), "left", "center", "right",
	ymd("1985-01-01"), 13, ymd("1979-01-01"), 15, balance_lines("in the 40 years between 1960 and 2000, the homicide rate trippled", 3), "right", "center", "right",
	ymd("2002-06-01"), 19.7, ymd("2004-06-01"), 18.7, balance_lines("the 173 Harold Shipman murders are recorded as all occurring in 2002", 5), "left", "center", "right",
	ymd("2013-06-01"), 9.1, ymd("2008-06-01"), 8.1, balance_lines("after 15 years of decreases, the homicide rate has been rising since 2014", 3), "right", "center", "left"
	# ymd("1961-06-01"), 5.5, ymd("1966-06-01"), 4.5, balance_lines("", 2), "left", "center", "right"
)
```

```{r build plot}
chart <- ggplot(chart_data, aes(year, homicide_rate)) +
  geom_point(aes(alpha = "point"), colour = "grey75", na.rm = TRUE) +
  geom_smooth(aes(colour = "line"), method = "loess", formula = "y ~ x", span = 0.2, se = FALSE, 
              na.rm = TRUE, key_glyph = "timeseries", alpha = 1) +
	# add explanatory labels
	geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "right"), inherit.aes = FALSE, 
						 curvature = elements$label_line_curvature, 
						 colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_segment(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "straight"), 
						 inherit.aes = FALSE, colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_curve(aes(x = x, y = y, xend = xend, yend = yend),
						 data = filter(chart_labels, curve == "left"), inherit.aes = FALSE, 
						 curvature = elements$label_line_curvature * -1, 
						 colour = elements$label_line_colour, 
						 arrow = elements$label_arrow, show.legend = FALSE) +
	geom_label(aes(x = xend, y = yend, label = label, hjust = hjust, 
								 vjust = vjust),
						data = chart_labels, inherit.aes = FALSE, 
						colour = elements$label_text_colour,
						fill = elements$label_text_fill, size = elements$label_text_size, 
						lineheight = elements$label_text_lineheight,
						label.size = NA, show.legend = FALSE) +
	# end of explanatory labels
  scale_x_date(date_breaks = "10 years", date_labels = "%Y", 
               expand = expand_scale(mult = 0.025)) +
  scale_y_continuous(limits = c(0, NA), 
                     expand = expand_scale(mult = c(0, 0.05))) +
  scale_colour_manual(
    values = c("line" = ucl_colours_list[["Orange"]]),
    labels = c("line" = "trend estimated from annual rates")
  ) + 
  scale_alpha_manual(
    values = c("point" = 1),
    labels = c("point" = "annual homicide rate")
  ) +
	labs(
		title = chart_details$title,
		subtitle = format_subtitle(chart_details$subtitle),
		legend = NULL,
		x = NULL,
		y = "homicides per million people",
		alpha = NULL,
		colour = NULL
	) +
	theme_cjcharts() +
  theme(
    legend.background = element_rect(fill = "white", colour = "white"),
    legend.direction = "horizontal",
    legend.justification = c(0, 0),
    legend.key.width = unit(1, "lines"),
    legend.margin = margin(3, 4, 3, 4),
    legend.position = c(0, 0.025),
    legend.spacing = unit(0, "lines"),
    legend.text = element_text(colour = elements$label_text_colour)
  )
```

`r chart_details$subtitle`

```{r display plot, echo=FALSE, include=TRUE}
add_logo(chart + labs(title = NULL, subtitle = NULL), 
				 chart_details$source_title, chart_details$id)
```

[larger image](../`r chart_details$id`.png)
| [annotated R code to produce this chart](https://github.com/mpjashby/lesscrime.info/blob/master/content/post/`r chart_details$id`.Rmd)

Data sources: [Homicide counts 1898 to 2001/02](`r chart_details$source_url`),
[Homicide counts 2001/02 onwards](https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/crimeinenglandandwalesappendixtables), [Population estimates 1838 to 2018](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland).

### Notes

This chart uses police-recorded homicide offences because that time series 
extends further back in time than the [Home Office Homicide Index](https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/articles/homicideinenglandandwales/yearendingmarch2018#how-is-homicide-defined-and-measured), the main 
alternative source of homicide data. The main disadvantage of this approach is 
that crimes that are initially recorded homicides in police data but are 
subsequently discovered not to be will not be included in the Homicide Index
data but will be included in police-recorded counts.

Victims of the Hillsborough disaster are recorded in ONS data as manslaughter 
cases in the year ending March 2017 because that was when the revised inquest 
verdicts were returned. However, in the chart above these cases are included in
the rate for 1989, when the deaths occurred.

```{r export chart}
# save PNG for social media
ggsave(
	filename = paste0(chart_details$id, ".png"), 
	plot = add_logo(chart, chart_details$source_title, chart_details$id), 
	device = "png", width = 600 / 72, height = 400 / 72, units = "in"
)
```
