---
title: "Chart: Typical rape investigation times doubled between 2012 and 2018"
author: Matt Ashby
date: '2021-03-17'
categories:
  - Crime and justice chart of the week
tags:
  - criminal justice
  - courts
  - prosecutions
  - trials
---
	
```{r set knitr options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, echo=FALSE, include=FALSE)
```

```{r set chart parameters}
chart_details <- list(
	id = "rape-prosecution-times",
	title = "Typical rape investigation times doubled between 2012 and 2018",
	subtitle = "The time taken for a typical rape case to be prosecuted in court increased only slightly from 2012 to 2018, but the time taken for police to investigate rapes and the CPS to decide whether to charge a suspect has more than doubled. The government has not published updated data since March 2019.",
	source_url = "https://www.gov.uk/government/statistics/criminal-court-statistics-quarterly-july-to-september-2019",
	source_title = "Ministry of Justice, 2019"
)
```

```{r load packages and helper}
# custom packages not loaded by helpers.R

# load this after loading custom packages
source(here::here("helpers.R"))
```

```{r get and tidy data}
# get data
if (!file.exists(paste0(chart_details$id, "-data.csv.gz"))) {
	
	# the data in the file https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/851931/time_offence_tool.xlsx are hidden in a pivot table, so we load a version manually extracted from the pivot table, including data for the most-recent available year
	
	tidy_data <- read_csv("rape-prosecution-times-raw-data.csv") %>% 
		clean_names() %>% 
		rename(year = row_labels)
	
	# save tidy data
	write_csv(tidy_data, paste0(chart_details$id, "-data.csv.gz"))	
	
} else {
	
	# load tidy data
	tidy_data <- read_csv(paste0(chart_details$id, "-data.csv.gz"))
	
}
```

```{r prepare data for chart}
chart_data <- tidy_data %>% 
	mutate(
		offence_to_charge = median_offence_to_charge_days,
		offence_to_first_hearing = median_charge_to_first_listing_days,
		offence_to_main_hearing = 
			median_charge_to_first_listing_days +
			median_sending_to_crown_court_to_main_hearing_days,
		offence_to_completion = 
			median_charge_to_first_listing_days +
			median_sending_to_crown_court_to_main_hearing_days +
			median_main_hearing_to_completion_days
	) %>% 
	select(year, offence_to_charge, offence_to_first_hearing, 
				 offence_to_main_hearing, offence_to_completion) %>% 
	mutate(
	) %>% 
	pivot_longer(-year, names_to = "period", values_to = "days") %>% 
	filter(!is.na(year)) %>% 
	mutate(
		months = days / (365 / 12),
		period = recode_factor(
			period,
			"offence_to_charge" = "offence to suspect being charged",
			"offence_to_first_hearing" = 
				"suspect being charged to first court appearance",
			"offence_to_main_hearing" = 
				"first court appearance to main hearing",
			"offence_to_completion" = 
				"main hearing to case being completed"
		),
		phase = ifelse(
		  period == "offence to suspect being charged",
		  "investigation by police and charging decision by CPS",
		  "prosecution in the courts"
		),
		year = fct_reorder(as.character(year), -year)
	) %>% 
	arrange(year, period) %>% 
	group_by(year, phase) %>% 
	mutate(prev = ifelse(is.na(lag(months)), 0, lag(months))) %>% 
	ungroup()
```

```{r build plot}
chart <- ggplot(
	chart_data, 
	aes(x = prev, y = year, xend = months, yend = year, colour = period)
) +
	geom_segment(
		size = 1.25, 
		arrow = arrow(angle = 40, length = unit(6, "points"), type = "closed")
	) +
  facet_grid(cols = vars(phase), scales = "free_x", space = "free_x") +
	scale_x_continuous(
		breaks = seq(0, 24, by = 3),
		minor_breaks = seq(0, 24, by = 1),
		expand = expansion(mult = c(0, 0.02))
	) +
	scale_colour_manual(
		guide = guide_legend(nrow = 2),
		values = unname(ucl_colours_list[c("Mid Blue", "Bright Blue", "Light Blue", 
																			 "Blue Celeste")])
	) +
	labs(
		title = chart_details$title,
		subtitle = format_subtitle(chart_details$subtitle, .width = 105),
		legend = NULL,
		x = "median months from rape being committed/defendant being charged, Crown Court cases only",
		y = NULL,
		colour = NULL
	) +
	theme_cjcharts() +
	theme(
		legend.box.margin = margin(0, 0, 0, 0),
		legend.direction = "vertical",
		legend.margin = margin(0, 0, 0, 0),
		panel.grid.major.x = element_line(),
		panel.grid.minor.x = element_line(),
		panel.grid.major.y = element_blank()
	)
```

`r chart_details$subtitle`

```{r display plot, echo=FALSE, include=TRUE}
add_logo(chart + labs(title = NULL, subtitle = NULL), 
				 chart_details$source_title, chart_details$id)
```

There are several potential reasons for this increase, which could explain it
either individually or in combination:

  * Cuts to police resources may have led to longer investigation times.
  * Cuts to CPS resources may have led to longer delays in making the decision
    whether or not to charge the suspect.
  * Attempts to increase the proportion of rapes reported to police may mean 
    more rape investigations need extended inquiries.
  * Developments in investigative techniques (e.g. analysis of mobile-phone
    data) or changes in offence circumstances may mean investigations take 
    longer.

[larger image](../`r chart_details$id`.png)
| [annotated R code to produce this chart](https://github.com/mpjashby/lesscrime.info/blob/master/content/post/`r chart_details$id`.Rmd)

Data source: [`r chart_details$source_title`](`r chart_details$source_url`)

```{r export chart}
# save PNG for social media
ggsave(
	filename = paste0(chart_details$id, ".png"), 
	plot = add_logo(chart, chart_details$source_title, chart_details$id), 
	device = "png", width = 600 / 72, height = 400 / 72, units = "in"
)
```
